Temporal dataflow graphs and their application to generative music

### 2017-03-21 article-icmc2017.tex
save work

\usepackage{icmc2017template}
\usepackage{times}
\usepackage{ifpdf}
\usepackage{soul}
\usepackage{microtype}
\usepackage{polyglossia}
%\usepackage{geometry}

@@

	\section{Implicit and explicit connections}
	Inner data : reduces the possibilities of connection : everything has to be there (explicit)

	Outer data : have to care about overwriting (implicit)
    Implicit edges allow for features such as pattern-matching : one node outputs to random, unknown other inputs, or to 
    the global state.
    For inputs to handle this, we have to provide them with a "can accept" function.

	\section{Relationships}
    \subsection{Strict}
    An execution of a sub-graph can only happen when all the nodes of the sub-graph are executing.

	\subsection{Glutton}
	An execution of a sub-graph can happen whenever any of the nodes of a sub-graph are executing. 
	An order has to be provided.

@@

	\subsection{Parallelized}
	All the sub-graphs are run with the same input environment. 
	The resulting values put in the environment are undefined, but it is guaranteed that all values returned will have been set.

@@

	\subsection{Replaced}
	All the sub-graphs are run with the same input environment. 
	The resulting values put in the environment are those of the latest graph execution.

@@

	\section{Meaning}
	Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
	\section{Implementation}
    \section{Conclusion}
    Additional consideration : see PLOP 2006 article (multi-stream, etc), sample accuracy, etc

>>>>

\usepackage{icmc2017template}
\usepackage{times}
\usepackage{ifpdf}
\usepackage{soul}
\usepackage{microtype}
\usepackage{polyglossia}
%\usepackage{geometry}
\usepackage{tikz}
\usepackage{listings}

@@

    Throughout this paper, we will use the execution trace in fig.~\ref{fig.simple} as running example. 
    \begin{figure}
        \centering
        \begin{tikzpicture}[scale=0.5]
        \draw[line width=2pt] (0, 0) -- (0, 2) -- (5, 2) -- (5, 0) -- (0, 0);
        \draw[line width=2pt] (0 + 3, 0 - 2.5) -- (0 + 3, 2 - 2.5) -- (5 + 3, 2 - 2.5) -- (5 + 3, 0 - 2.5) -- (0 + 3, 0 - 2.5);

        \draw[line width=2pt] 
        (0, -3.75) -- 
        (0, -4) -- (3, -4) -- 
        (3, -3.75) --
        (3, -4) -- (5, -4) -- 
        (5, -3.75) -- 
        (5, -4) -- (8, -4) -- 
        (8, -3.75);

        \node (t1) at (1.45, -3.45) {\large $t_1$};
        \node (t2) at (4, -3.45) {\large $t_2$};
        \node (t3) at (6.5, -3.45) {\large $t_3$};

        \node (f1) at (2.5, 1) {\large $f_1$} ;
        \node (f2) at (5.5, -1.5) {\large $f_2$} ;
        \end{tikzpicture}
        \caption{A possible trace of execution for two processes}
        \label{fig.simple}
    \end{figure}

	\section{Environment, implicit and explicit connections}
    In the traditional data flow paradigm, nodes are linked together through connections in their respective input and output ports.

    We extend this with a permanent environment, that contains a mapping of key-value pairs. 
    Keys are adresses : \lstinline|/foo/bar| and values can be any kind of data : numbers, strings, etc. 
    The values of the environment may vary independently of the execution of the program : for instance elements of the environment could be changed through the network, graphical or physical controls.

    We then have to provide a mean to use the values of this environment to the dataflow graph.
    Thus, we extend the inputs and outputs ports with the following expliciteness semantic :

    \begin{itemize}
        \item An explicit port has been connected by the developer to another port.
        \item An implicit port has not been connected to another port but may accept any number of adress - value pairs from the environment to perform its work.
    \end{itemize}

    During a tick, adresses have access to two sets of values : 
    \begin{itemize}
        \item The values that were in the environment at the beginning of the tick.
        \item The values produced by previous nodes in the graph.
    \end{itemize}

    It is up to the developer of nodes of the graph to choose whether a port can only work when connected explicitely, or with the environment produced from previous nodes, or with the global environment, or both.

    This mechanism allows for new patterns during the creation of a dataflow program : for instance, instead of accepting data from a single source, a port could specify a pattern (see OSC 1.1 paper with XPath) such as : \lstinline|/foo.*/bar/b[a-zA-Z]|. 
    % TODO pattern-matched addresses on the output : seems hard ? how to match it with a "pattern-matched" input port ??
	%Inner data : reduces the possibilities of connection : everything has to be there (explicit)

	%Outer data : have to care about overwriting (implicit)
    %Implicit edges allow for features such as pattern-matching : one node outputs to random, unknown other inputs, or to the global state.
    % For inputs to handle this, we have to provide them with a "can accept" function.

    % TODO sample-precision of the activation / deactivation of a process / node
    % TODO 
	\section{Relationships}
    We are interested in the relationships between nodes of the dataflow graph when they produce compatible tokens.

    \subsection{Strict}
    An execution of a sub-graph can only happen when all the nodes of the sub-graph are executing.

    That is, if two nodes are compatible, the output of the first node can only go in the input of the second node when both are active.

    For instance, take the case in fig.~\ref{fig.simple} where $f_1$ and $f_2$ both read from $/a$ and write to $/a$.

    Let $t$ be the time of the tick and $e$ the global environment. We assume that $f_1, f_2$ always have access to this information.

    We will get the following set of equations representing the behaviour : 
    \begin{itemize}
        \item During $t_0$ : $\emptyset$.
        \item During $t_1$ : $f_2 \circ f_1 $ or $f_1 \circ f_2$.
        \item During $t_2$ : $\emptyset$. 
    \end{itemize}

    The strict execution may happen explicitely or implicitely.

	\subsection{Glutton}
	An execution of a sub-graph can happen whenever any of the nodes of a sub-graph are executing.

    If we take the same case than previously, the behaviour is :
    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_2 \circ f_1$ or $f_1 \circ f_2$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

    The glutton execution may happen explicitely or implicitely.

@@

    Note : this question has an answer in PLOP2006.

    The delayed execution has be specifieds explicitely for the tokens to be bufferised.

	\subsection{Parallelized and mixed}
	All the sub-graphs are run with the same input environment. 
	The resulting values put in the environment are undefined, but it is guaranteed that all values returned will have been set, which may trigger event handlers.

@@

    Thus, we have to define summing operations for each of the data types supported.

    We propose for instance : 
    \begin{itemize}
        \item For simple parameters such as MIDI control changes or OSC messages : the latest value replaces the previous ones.
        \item For audio streams : the streams are mixed together.
        \item For MIDI notes : the notes are played together.
    \end{itemize}

    The parallelized execution may happen explicitely or implicitely.

    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_1$ and $f_2$ run in parallel, with the same value for $/a$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

	\subsection{Replaced}
	All the nodes are run with the same input environment. 
	The resulting values put in the environment are those of the latest graph execution.

    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_1$ or $f_2$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

    The replaced execution may happen explicitely or implicitely.

@@

    That is, if the execution happens in the order of the fig.~\ref{fig.simple}, we get the following execution :

    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_2 \circ f_1$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

Instead, if we are in the case of fig.~\ref{fig.simple-reversed}, we get for $t_1$ the execution $f_1 \circ f_2$.

    \begin{figure}
        \centering
        \begin{tikzpicture}[scale=0.5]
        \draw[line width=2pt] (0 + 3, 0) -- (0 + 3, 2) -- (5 + 3, 2) -- (5 + 3, 0) -- (0 + 3, 0);
        \draw[line width=2pt] (0, 0 - 2.5) -- (0, 2 - 2.5) -- (5, 2 - 2.5) -- (5, 0 - 2.5) -- (0, 0 - 2.5);

        \draw[line width=2pt] 
        (0, -3.75) -- 
        (0, -4) -- (3, -4) -- 
        (3, -3.75) --
        (3, -4) -- (5, -4) -- 
        (5, -3.75) -- 
        (5, -4) -- (8, -4) -- 
        (8, -3.75);

        \node (t1) at (1.45, -3.45) {\large $t_1$};
        \node (t2) at (4, -3.45) {\large $t_2$};
        \node (t3) at (6.5, -3.45) {\large $t_3$};

        \node (f1) at (5.5, 1) {\large $f_1$} ;
        \node (f2) at (2.5, -1.5) {\large $f_2$} ;
        \end{tikzpicture}
        \caption{A possible trace of execution for two processes}
        \label{fig.simple-reversed}
    \end{figure}

	\section{Evaluation}
    We present here the evaluation algorithm that operates at each tick.

	Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
	\section{Implementation}
    \section{Conclusion}
    Additional consideration : see PLOP 2006 article (multi-stream, etc), sample accuracy, etc
